---
title: "Titanic Survival"
author: "Hana Le"
date: "2023-04-05"
output: 
 html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Introduction

This project uses data from the Kaggle competition ["Titanic - Machine Learning from Disater"](https://www.kaggle.com/competitions/titanic/overview). The goal of this project is to use machine learning to build a predictive model that predicts which passengers survived the Titanic shipwreck, to  answers the question: “what sorts of people were more likely to survive?” using passenger data (ie. name, age, gender, socio-economic class, etc). 

It seems like a straightforward project that is quite popular among beginners in data science :smiley:, but my objective is to gain more experience working with this type of dataset.

# 2. Overview the data
## 2.1 Loading packages and reading the data

```{r housing, message = FALSE, results='hide'}
# Loading R packages
packages <- c("tidyverse","Amelia", "psych","DT","mice","ranger", "janitor", "stringr","ggplot2", "GGally", "randomForest") 
sapply(packages, require, character = TRUE)
```


```{r}
# Reading data
train <- read.csv("titanic_data/train.csv")
test <- read.csv("titanic_data/test.csv")
```


## 2.2 Data size
The Titanic train data set has 891 obs and 12 variables with the response variable Survived. The Titanic test data set has 418 obs and 11 variables.

```{r}
dim(train); dim(test)
```

```{r combine 2 datasets}
# Combine 2 data sets to see the structure, and for cleaning & feature engineering later.
test$Survived <- NA
data <- rbind(train, test)
dim(data) 
```
The data now has 12 columns consisting of 11 predictors and the response variable Survived.

## 2.3 Data Structure

```{r}
str(data)
```

Some categorical variables should be in factor form:

```{r}
var_cat <- c("Survived", "Pclass","Sex", "Embarked")
data[, var_cat] <- data.frame(lapply(data[, var_cat], as.factor))
```


## 2.3 Missingness of the data

It seems from the data structure that missing values are not only represented as 'NA', but also as empty.

```{r}
sort(colSums(is.na(data)| data == ""), decreasing = T)
```

Out of a total of 1309 records, Cabin is the variable with the highest number of missing values, with 1014 records (77.5%) missing. Age is the second variable with the most missing values, with 263 records (20.1%) missing. Embarked has only 2 missing values, while Fare has only 1 missing value.

```{r}
data[data ==""] <- NA

missmap(data[,colnames(data) != "Survived"])
```


## 2.4 Imputing missing data


```{r}
# Imputing missing values for Embarked
data[is.na(data$Embarked), ]
data %>% tabyl(Embarked, Pclass)
#names(sort(table(data$Embarked), decreasing = T))[1]
# These 2 passengers were in 1st class, in the same Cabin
# It seems likely the 2 passengers embarked from S = Southampton.
data$Embarked[is.na(data$Embarked)] <- "S"
```

```{r}
# Imputing the missing value for Fare
data[is.na(data$Fare),]
## the passenger was in Pclass 3 and embarked from "S"
data$Fare[is.na(data$Fare)] <- median(data$Fare, data$Pclass == 3 & data$Embarked == "S", na.rm = T)
```


```{r}
# I was wondering if there is any case where Cabin is exclusively for the 1st class passengers?
# Have look on Cabin

# data %>% tabyl(Pclass, Cabin)

# It appears that passengers who had cabin information were predominantly in the 1st class. 
# However, there were still 67 1st-class passengers who had missing values for the cabin,
# making it difficult to draw conclusions. 
# I decided to drop the cabin variable from the analysis

data$Cabin <- NULL
```


```{r, warning=FALSE}
# Imputing missing values for Age
# I'm going to use mice package for Age
# select only the columns needed for imputing Age
age_data <- data[, c('Age', 'Pclass', 'Sex', 'SibSp', 'Parch', 'Fare', 'Embarked')]

# impute missing values using MICE with the "random forest" method

age_mice <- mice(age_data, method = "rf")
age_imputed <- complete(age_mice)
data$Age <- age_imputed$Age
```


```{r}
# Checking missing values
colSums(is.na(data[,colnames(data) != "Survived"]))
```

## 2.5 Descriptive statistics


```{r}
data_table <- describe(data)
data_table %>% round(digits = 3) %>%
  DT::datatable(options = list(pageLength = 10)) 
```

## 3. Exploring variables
### 3.1 Survived

```{r}
tabyl(data$Survived)
```
Among the 1309 passengers in the Titanic dataset, 342 survived and a target of 418 survival outcomes was set for the test set.



## 3. Feature Engineering
Passenger titles provide some information about their social status, gender, and possibly age. These titles were included in the name of each passenger so we can extract them out.

```{r}
data$Title <- str_extract(data$Name, "(?<=\\,\\s)[A-Za-z]+")
#data$Title <- sapply(data$Name, function(x) strsplit(x, split = '[,.]')[[1]][[2]])
#data$Title <- sub(' ', '', data$Title)
data %>% tabyl(Title, Sex) 
```

```{r}
data$Title[data$Title %in% c("Mlle", "Ms")] <- "Miss"
data$Title[data$Title == "Mme"] <- "Mrs"
data$Title[data$Title %in% c("the Countess", "Dona", "Lady", "Jonkheer")] <- "Noble"
data$Title[data$Title %in% c("Capt","Col", "Dr", "Rev", "Don", "Major", "Sir")] <- "Officer"
data %>% tabyl(Title, Sex) 

```

```{r}


```

### 4.1 Exploring predictors of Survived

### 4.1.1 Finding important predictors

```{r}
# Split data into training and testing sets
set.seed(1234)
train_index <- sample(nrow(titanic), round(nrow(titanic)*0.7), replace = FALSE)
train_data <- titanic[train_index, ]
test_data <- titanic[-train_index, ]

# Train random forest model
rf_model <- randomForest(Survived ~ ., data = train_data, ntree = 500, importance = TRUE)

# Compute variable importance measures
vi <- importance(rf_model)

# Create a data frame with the variable names and importance measures
vi_df <- data.frame(variable = names(vi), importance = vi)

# Sort the data frame by importance measures in descending order
vi_df <- vi_df[order(-vi_df$importance),]

# Create a barplot with ggplot
library(ggplot2)
ggplot(vi_df[1:10,], aes(x = reorder(variable, importance), y = importance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "Variables", y = "Importance", title = "Variable Importance Plot") +
  theme_bw() +
  coord_flip()
```



### 4.1.2 Visualising relation of Survided with the most important variables.

```{r}
# Subset the dataset to include only the variables of interest
set_con <- subset(data[!is.na(data$Survived),], select = c("Survived", "Age", "Fare"))

# Create scatter plot matrix with ggally
ggpairs(sub_con, aes(color = Survived, alpha = 0.7),
        lower = list(combo = "count"))
```

The median age of the survivors appears to be slightly lower than that of non-survivors, indicating that younger individuals had a higher chance of survival. On the other hand, the median fare of the surviving passengers is noticeably higher, suggesting a potential correlation between fare paid and survival rate.

## 5. Data preparation for modelling
## 6. Modelling
## 7. Conclusion
